---
title: "Final Project"
author: "Logan Parker"
date: "11/14/2020"
output: html_document
---

```{r setup, include=FALSE,echo=FALSE,message=FALSE}
knitr::opts_chunk$set(echo = FALSE)
#loading in all of the necessary r packages
library(tidyverse)
library(leaflet)
library(lubridate)
library(usmap)
library(cowplot)
library(geojsonio)
library(htmltools)
```

## Question 1
### Can the number of prisoners in a state be used to predict how safe the state is?

The following graphs show the prisoner count by state and then the prisoner count adjusted for population.
The graph for raw number of prisoners in the state looks almost identical to the graph that shows the raw number of 
violent crimes in that state. Further exploring this connection with a scatter plot we can see that the number of prisoners in a state
is almost perfectly coorrelated to the number of violent crimes.
```{r prisoner,message=FALSE}
#Loading in the crime dataset and wrangling the states to match the us_map package
crime <- read_csv("data/crime_and_incarceration_by_state.csv")
crime <- crime %>%
  rename("state" = jurisdiction) %>%
  mutate(state = tolower(state))

#creating a statemap and renaming the states to match the above crime dataset
state_map <- us_map("states") %>%
  rename(state = full) %>%
  mutate(state = tolower(state))
#Here I am filtering the dataset to only include 2016, the newest, data and creating a prisoner_map with only state, prisoner count, population, and crime by population and then joining the crime and statemap on the previously adjusted state variable.
prisoner_map <- crime %>%
  filter(year == 2016, state != "federal") %>%
  group_by(state,prisoner_count,state_population)%>%
  summarise(crimePop = prisoner_count/state_population)%>%
  left_join(state_map,.,on = "state")
#This section is where the mapping of prison_map takes place. The map is filled by basic prisoner count on a scale from yellow to red
prisoner_map %>%
  ggplot(mapping = aes(x = x,
                       y = y,
                       group = group,
                       fill = prisoner_count))+
  geom_polygon(color = "gray") +
  coord_equal() +
  theme_map()+
  scale_fill_distiller(palette = "YlOrRd",direction = 1)+
  labs(title = "Number of Prisoners by State 2016",fill = "Prisoner Count")
#This section maps prison_map again, but this time the map is filled by prisoner count adjusted for population of the state on a scale from yellow to red
prisoner_map %>%
  ggplot(mapping = aes(x = x,
                       y = y,
                       group = group,
                       fill = crimePop))+
  geom_polygon(color = "gray") +
  coord_equal() +
  theme_map()+
  scale_fill_distiller(palette = "YlOrRd",direction = 1)+
  labs(title = "Prisoners vs Population 2016",fill = element_blank())

```

## Question 2
### Are the number of violent crimes correlated to the total population of an area?

The following graphs show the raw number of violent crimes in a state and the number of violent crimes per capita. It can be seen
that the number of violent crimes is very closely coorrelated to the total population of an area.
```{r violent, message=FALSE}
#Here I am creating a map datatable that includes map data as well as population and violent crime total
violent_map <- crime %>%
  filter(year == 2016, state != "federal") %>%
  group_by(state,state_population,violent_crime_total)%>%
  summarise(violentPop = violent_crime_total/state_population)%>%
  left_join(state_map,.,on = "state")
#The plotting of this data table where I just fill by violent crime total 
violent_map %>%
  ggplot(mapping = aes(x = x,
                       y = y,
                       group = group,
                       fill = violent_crime_total))+
  geom_polygon(color = "gray") +
  coord_equal() +
  theme_map()+
  scale_fill_distiller(palette = "YlOrRd",direction = 1)+
  labs(title = "Violent Crime by State 2016",fill = "Violent Crimes")
#This plot is filled using the violent crime adjusted for population
violent_map %>%
  ggplot(mapping = aes(x = x,
                       y = y,
                       group = group,
                       fill = violentPop))+
  geom_polygon(color = "gray") +
  coord_equal() +
  theme_map()+
  scale_fill_distiller(palette = "YlOrRd",direction = 1)+
  labs(title = "Violent Crimes vs Population 2016",fill = element_blank())
#Here, I am creating a subset of crime that only includes 2016 data
prisoners16 <- crime %>%
  filter(year == 2016, state != "federal")
#This is the scatter plot that shows the coorrelation coefficient in the title for population and violent crime
prisoners16%>%
  ggplot(mapping = aes(state_population,violent_crime_total))+
  geom_point()+
  labs(x="Population",y="Violent Crimes",title = paste("Coorelation of population and Violent Crime:",cor(prisoners16$state_population,prisoners16$violent_crime_total)))

#This is the scatter plot that shows the coorrelation coefficient in the title for prisoner count and violent crime
prisoners16%>%
  ggplot(mapping = aes(prisoner_count,violent_crime_total))+
  geom_point()+
  labs(x="Prisoner Count",y="Violent Crimes",title = paste("Coorelation of Prisoner Count and Violent Crime:",cor(prisoners16$prisoner_count,prisoners16$violent_crime_total)))

prisoners16%>%
  ggplot(mapping = aes(prisoner_count,violent_crime_total/state_population))+
  geom_point()+
  labs(x="Prisoner Count",y="Violent Crimes per capita",title = paste("Coorelation of Prisoner Count and Violent Crime per capita:",cor(prisoners16$prisoner_count,(prisoners16$violent_crime_total/prisoners16$state_population))))

prisoners16%>%
  ggplot(mapping = aes(state_population,violent_crime_total/state_population))+
  geom_point()+
  labs(x="Population",y="Violent Crimes per capita",title = paste("Coorelation of population and Violent Crime per capita:",cor(prisoners16$state_population,(prisoners16$violent_crime_total/prisoners16$state_population))))

```

## Bar

```{r bar, message=FALSE}
#Here I am creating a barplot to show the number of prisoners in each state and reordering it to be in descending order
prisoners16 %>%
  ggplot(mapping = aes(x=reorder(state,prisoner_count),y=prisoner_count, fill=state))+
  geom_bar(stat = "identity")+
  coord_flip()+
  guides(fill=FALSE)+
  labs(title = "Number of Prisoners by State in 2016",x=element_blank(),y="Prisoner Count")

```

## Summary of the information

```{r leaflet, message=FALSE}
#this geojson file has the outlines of the states needed to make the highlight funcion work properly
states <- geojsonio::geojson_read("data/us-states.json",what = "sp")
#this line creates a variable that represents the number of prisoners per 1000 citizens
prisoners16 <- prisoners16 %>%
  mutate(prisonPop = (prisoner_count/state_population *1000),
         violentPop = (violent_crime_total/state_population * 1000))
#Here I generate labels for the states when they are moused over
labels <- sprintf(
  "<strong>%s</strong><br/>%d total prisoners<br />Total population: %d <br />Prisoners per 1000 Citizens: %g<br />Violent Crimes per 1000 Citizens: %g",
  prisoners16$state, prisoners16$prisoner_count, prisoners16$state_population, prisoners16$prisonPop, prisoners16$violentPop
) %>% lapply(htmltools::HTML)
#this line defines a custom pallet which allows me to fill the polygons by prisonPop
pal <- colorBin("YlOrRd", domain = prisoners16$violentPop)
#Here is my leaflet that is interactive and shows the labels produced. My mapbox token is used to allow the better looking mapbox package to be used.
leaflet(states) %>%
  setView(-96, 37.8, 4) %>%
  addProviderTiles("MapBox", options = providerTileOptions(
    id = "mapbox.light",
    accessToken = "pk.eyJ1IjoibG9nYW5wYXJrMTIzIiwiYSI6ImNraGk3aDczeDA3YmwyeHM4bG44bG81cTAifQ.Y_t7mMxNVcc6TMuy5tfccw"))%>%
  addPolygons(
    fillColor = ~pal(prisoners16$violentPop),
    weight = 2,
    opacity = 1,
    color = "white",
    dashArray = "3",
    fillOpacity = 0.7,
    highlight = highlightOptions(
      weight = 5,
      color = "#666",
      dashArray = "",
      fillOpacity = 0.7,
      bringToFront = TRUE),
    label = labels,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal", padding = "3px 8px"),
      textsize = "15px",
      direction = "auto")
    ) %>%
  addLegend(pal = pal, values = ~prisoners16$violentPop, opacity = 0.7, title = "Violent Crimes per 1000 citizens",
            position = "bottomright")
```